Evaluation of classical correlation functions from 2/3D images on CPU and GPU architectures: introducing CorrelationFunctions.jl

### 2024-01-31 paper-correlations.tex
Enlarge figure 4
added final changes
Prepare to publishing

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.9\linewidth]{images/cf_1.png}
  \caption[]{A schematic representation of a binary porous media (pores are
    shaded) with examples of CFs computations (all classical correlation
    functions are presented according to the legend on the right
    side). Computations take into account all possible configurations of a line
    segment (or a N-dimensional ball for pore size function) which either make a
    contribution to the CF (conventionally shown as ``CF = 1'') or do not make a
    contribution (``CF = 0'').}
  \label{fig:functions}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.9\linewidth]{images/cf_diag.png}
  \caption[]{Available directions to compute directional correlation functions
    in 2D and 3D cases with line scanning approach.}
  \label{fig:directions}
\end{figure}

@@

\subsection{Comparison against analytical solutions}
To test out package we generate a realization of random overlapping disks
(\cref{fig:overlapping-disks}) with dimensions $10000 \times 10000$ and
parameters of radii $R = 40\ pixels$ and $\lambda = 10^{-4}\ pixels^{-2}$ and
then compare calculated correlation functions with their theoretical values.
\begin{figure}[ht]
  \centering
  \includegraphics[width=0.45\linewidth]{images/disks.png}
  \includegraphics[width=0.45\linewidth]{images/balls.png}
  \caption[]{Images of overlapping disks and balls used for testing of our package.}
  \label{fig:overlapping-disks}
\end{figure}

@@

\begin{figure*}[tp]
  \centering
  \subfigure[Lineal-path correlation function]{
    \includegraphics[width=0.45\linewidth]{images/l2.png}
    \label{fig:l2}}
  \hfill
  \subfigure[Two-point correlation function]{
    \includegraphics[width=0.45\linewidth]{images/s2.png}
    \label{fig:s2}}
  \vskip\baselineskip
  \subfigure[Surface-surface correlation function]{
    \includegraphics[width=0.45\linewidth]{images/ss.png}
    \label{fig:ss}}
  \hfill
  \subfigure[Surface-void correlation function]{
    \includegraphics[width=0.45\linewidth]{images/sv.png}
    \label{fig:sv}}
  \vskip\baselineskip
  \subfigure[Pore size correlation function]{
    \includegraphics[width=0.45\linewidth]{images/ps.png}
    \label{fig:ps}}
  \hfill
  \subfigure[Chord length correlation function]{
    \includegraphics[width=0.45\linewidth]{images/cl.png}
    \label{fig:cl}}
  \caption[]{A comparison of calculated and theoretical values of correlation
    functions for overlapping disks and balls.}
  \label{fig:verification}
\end{figure*}

@@

\begin{figure*}[pt]
  \centering
  \subfigure[Ceramic 1]{
    \includegraphics[width=0.3\linewidth]{samples/Ceramic1.png}}
  \hfill
  \subfigure[Ceramic 2]{
    \includegraphics[width=0.3\linewidth]{samples/Ceramic2.png}}
  \hfill
  \subfigure[Ceramic 3]{
    \includegraphics[width=0.3\linewidth]{samples/Ceramic3.png}}
  \vskip\baselineskip
  \subfigure[Sandstone 1]{
    \includegraphics[width=0.3\linewidth, frame]{samples/sandstone1.png}}
  \hfill
  \subfigure[Sandstone 2]{
    \includegraphics[width=0.3\linewidth, frame]{samples/sandstone2.png}}
  \hfill
  \subfigure[Carbonate 1]{
    \includegraphics[width=0.3\linewidth, frame]{samples/carbonate1.png}}
  \caption[]{Visualization of XCT (top) and SEM (bottom) images used for
    computation of correlation functions. Different clusters of the void phase
    on SEM samples are shown in color.}
  \label{fig:real-data}
\end{figure*}
\begin{figure*}[pt]
  \centering
  \includegraphics[width=0.45\linewidth]{samples/s2-xct.png}
  \hfill
  \includegraphics[width=0.45\linewidth]{samples/c2-xct.png}
  \vskip\baselineskip
  \includegraphics[width=0.45\linewidth]{samples/l2-xct.png}
  \vskip\baselineskip
  \includegraphics[width=0.45\linewidth]{samples/surf2-xct.png}
  \hfill
  \includegraphics[width=0.45\linewidth]{samples/surfvoid-xct.png}
  \vskip\baselineskip
  \includegraphics[width=0.45\linewidth]{samples/pore-size-xct.png}
  \hfill
  \includegraphics[width=0.45\linewidth]{samples/chord-length-xct.png}
  \caption[]{Plots of correlation functions for XCT samples on
    \cref{fig:real-data}.}
  \label{fig:real-xct-cfs}
\end{figure*}

The \cref{fig:real-data} (at the bottom) presents three 2D images of sandstones
and carbonate with computed correlation functions at \cref{fig:real-sem-cfs}. In
this example, unlike for quite isotropic ceramics, we evaluated CFs in the
directional manner. The $S_2$ graph immediately provides us with porosity
values. It also highlights severe anisotropy in sandstone SEM images. For 2D
images numerous patches of pores are not connected to each other and, thus,
$S_2$ and $C_2$ are very different. Cluster functions show how connectivity
decays with length, also we can easily observe if given pore structure
percolates in a given direction - none of the images percolates (we highlighted
the clusters to make this statement easily verifiable with the naked eye). Note
that $C_2$ and $L_2$ were computed with correlations lengths equal to image
dimensions explore connectivity -- for other CFs this leads to noised curves due
to smaller number of sampled points (for example, see correlation functions'
graphs in \cite{Euras2012}).  Lineal paths are close for sandstones (with lower
$L_2$ values for Sandstone 1 where large pores are occluded with clays or grain
fragments), but for carbonate with much smaller apertures of pores it is way
below $L_2$ for sandstones. Also, $L_2$ suggests that computations with
correlations lengths beyond where lineal path goes to zero are not very
informative. For very irregular pores on 2D SEM images, $P$ and $p$ functions
look much less similar compared to the ceramics case on
\cref{fig:real-xct-cfs}. For 2D images it is also visually easy to analyze full
CF maps, as presented on \cref{fig:real-sem-map} for Sandstone 1 sample. Full
correlation maps show expanded version of directional functions -- asymmetry is
immediately evidenced from $S_2$, $F_{ss}$ and $F_{sv}$ maps. Moreover, the map
for $C_2$ shows higher connectivity in y-direction as compared to x-axis. The
interpretation of full correlation maps is less straightforward, but anisotropy
patterns in directions different from those presented in \cref{fig:directions}
can be spotted.
\begin{figure*}[pt]
  \centering
  \includegraphics[width=0.45\linewidth]{samples/s2-sem.png}
  \hfill
  \includegraphics[width=0.45\linewidth]{samples/c2-sem.png}
  \vskip\baselineskip
  \includegraphics[width=0.45\linewidth]{samples/l2-sem.png}
  \vskip\baselineskip
  \includegraphics[width=0.45\linewidth]{samples/surf2-sem.png}
  \hfill
  \includegraphics[width=0.45\linewidth]{samples/surfvoid-sem.png}
  \vskip\baselineskip
  \includegraphics[width=0.45\linewidth]{samples/pore-size-sem.png}
  \hfill
  \includegraphics[width=0.45\linewidth]{samples/chord-length-sem.png}
  \caption[]{Plots of correlation functions for SEM samples on
    \cref{fig:real-data}}
  \label{fig:real-sem-cfs}
\end{figure*}

\begin{figure*}[pt]
  \centering
  \subfigure[$S_2$]{
    \includegraphics[width=0.45\linewidth]{samples/map-s2.png}}
  \hfill
  \subfigure[$C_2$]{
    \includegraphics[width=0.45\linewidth]{samples/map-c2.png}}
  \vskip\baselineskip
  \subfigure[$F_{ss}$]{
    \includegraphics[width=0.45\linewidth]{samples/map-surf2.png}}
  \hfill
  \subfigure[$F_{sv}$]{
    \includegraphics[width=0.45\linewidth]{samples/map-surfvoid.png}}
  \caption[]{Correlation maps for Sandstone 1.}
  \label{fig:real-sem-map}
\end{figure*}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.45\linewidth]{samples/multiphase.png}
  \includegraphics[width=0.45\linewidth]{samples/multiphase-cc.png}
  \caption[]{Cross-correlation functions for multiphase sample (soil) which has
    4 distinct phases, labeled as 1-4. Each function $\rho_{ij}$ is normalized
    by division by $\rho_{ii}(0) \rho_{jj}(0)$.}
  \label{fig:real-multiphase}
\end{figure}

@@

\begin{figure*}[t]
  \centering
  \subfigure[Two-dimensional case, module \code{Directional}]{
    \includegraphics[width=0.475\linewidth]{images/time-2d.png}
    \label{fig:timings-2d}}
  \hfill
  \subfigure[Three-dimensional case, module \code{Directional}]{
    \includegraphics[width=0.475\linewidth]{images/time-3d.png}
    \label{fig:timings-3d}}
  \vskip\baselineskip
  \subfigure[Two-dimensional case, module \code{Map}, CPU]{
    \includegraphics[width=0.475\linewidth]{images/time-2d-map.png}
    \label{fig:timings-2d-map}}
  \hfill
  \subfigure[Three-dimensional case, module \code{Map}, CPU]{
    \includegraphics[width=0.475\linewidth]{images/time-3d-map.png}
    \label{fig:timings-3d-map}}
  \vskip\baselineskip
  \subfigure[Two-dimensional case, module \code{Map}, GPU]{
    \includegraphics[width=0.475\linewidth]{images/time-2d-gpu.png}
    \label{fig:timings-2d-gpu}}
  \hfill
  \subfigure[Three-dimensional case, module \code{Map}, GPU]{
    \includegraphics[width=0.475\linewidth]{images/time-3d-gpu.png}
    \label{fig:timings-3d-gpu}}
  \caption[]{Execution times for different correlation functions.}
  \label{fig:timings}
\end{figure*}

@@

\section{Acknowledgments}
This research was supported by the Russian Science Foundation grant 19-72-10082.

Collaborative effort of the authors within the FaT iMP (Flow and Transport in
Media with Pores) research group (www.porenetwork.com) and used some of its
software.

>>>>

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.9\linewidth]{cf_1.png}
  \caption[]{A schematic representation of a binary porous media (pores are
    shaded) with examples of CFs computations (all classical correlation
    functions are presented according to the legend on the right
    side). Computations take into account all possible configurations of a line
    segment (or a N-dimensional ball for pore size function) which either make a
    contribution to the CF (conventionally shown as ``CF = 1'') or do not make a
    contribution (``CF = 0'').}
  \label{fig:functions}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.9\linewidth]{cf_diag.png}
  \caption[]{Available directions to compute directional correlation functions
    in 2D and 3D cases with line scanning approach.}
  \label{fig:directions}
\end{figure}

@@

\subsection{Comparison against analytical solutions}
To test out package we generate a realization of random overlapping disks
(\cref{fig:overlapping-disks}) with dimensions $10000 \times 10000$ and
parameters of radii $R = 40\ pixels$ and $\lambda = 10^{-4}\ pixels^{-2}$ and
then compare calculated correlation functions with their theoretical values.
\begin{figure}[ht]
  \centering
  \includegraphics[width=0.45\linewidth]{disks.png}
  \includegraphics[width=0.45\linewidth]{balls.png}
  \caption[]{Images of overlapping disks and balls used for testing of our package.}
  \label{fig:overlapping-disks}
\end{figure}

@@

\begin{figure*}[tp]
  \centering
  \includegraphics[width=0.45\linewidth]{l2.png}
  \hfill
  \includegraphics[width=0.45\linewidth]{s2.png}
  \vskip\baselineskip
  \includegraphics[width=0.45\linewidth]{ss.png}
  \hfill
  \includegraphics[width=0.45\linewidth]{sv.png}
  \vskip\baselineskip
  \includegraphics[width=0.45\linewidth]{ps.png}
  \hfill
  \includegraphics[width=0.45\linewidth]{cl.png}
  \caption[]{A comparison of calculated and theoretical values of correlation
    functions for overlapping disks and balls.}
  \label{fig:verification}
\end{figure*}

@@

\begin{figure*}[pt]
  \centering
  \subfigure[Ceramic 1]{
    \includegraphics[width=0.3\linewidth]{Ceramic1.png}}
  \hfill
  \subfigure[Ceramic 2]{
    \includegraphics[width=0.3\linewidth]{Ceramic2.png}}
  \hfill
  \subfigure[Ceramic 3]{
    \includegraphics[width=0.3\linewidth]{Ceramic3.png}}
  \vskip\baselineskip
  \subfigure[Sandstone 1]{
    \includegraphics[width=0.3\linewidth, frame]{sandstone1.png}}
  \hfill
  \subfigure[Sandstone 2]{
    \includegraphics[width=0.3\linewidth, frame]{sandstone2.png}}
  \hfill
  \subfigure[Carbonate 1]{
    \includegraphics[width=0.3\linewidth, frame]{carbonate1.png}}
  \caption[]{Visualization of XCT (top) and SEM (bottom) images used for
    computation of correlation functions. Different clusters of the void phase
    on SEM samples are shown in color.}
  \label{fig:real-data}
\end{figure*}
\begin{figure*}[pt]
  \centering
  \includegraphics[width=0.45\linewidth]{s2-xct.png}
  \hfill
  \includegraphics[width=0.45\linewidth]{c2-xct.png}
  \vskip\baselineskip
  \includegraphics[width=0.45\linewidth]{l2-xct.png}
  \vskip\baselineskip
  \includegraphics[width=0.45\linewidth]{surf2-xct.png}
  \hfill
  \includegraphics[width=0.45\linewidth]{surfvoid-xct.png}
  \vskip\baselineskip
  \includegraphics[width=0.45\linewidth]{pore-size-xct.png}
  \hfill
  \includegraphics[width=0.45\linewidth]{chord-length-xct.png}
  \caption[]{Plots of correlation functions for XCT samples on
    \cref{fig:real-data}.}
  \label{fig:real-xct-cfs}
\end{figure*}

The \cref{fig:real-data} (at the bottom) presents three 2D images of sandstones
and carbonate with computed correlation functions at \cref{fig:real-sem-cfs}. In
this example, unlike for quite isotropic ceramics, we evaluated CFs in the
directional manner. The $S_2$ graph immediately provides us with porosity
values. It also highlights severe anisotropy in sandstone SEM images. For 2D
images numerous patches of pores are not connected to each other and, thus,
$S_2$ and $C_2$ are very different. Cluster functions show how connectivity
decays with length, also we can easily observe if given pore structure
percolates in a given direction - none of the images percolates (we highlighted
the clusters to make this statement easily verifiable with the naked eye). Note
that $C_2$ and $L_2$ were computed with correlations lengths equal to image
dimensions explore connectivity -- for other CFs this leads to noised curves due
to smaller number of sampled points (for example, see correlation functions'
graphs in \cite{Euras2012}).  Lineal paths are close for sandstones (with lower
$L_2$ values for Sandstone 1 where large pores are occluded with clays or grain
fragments), but for carbonate with much smaller apertures of pores it is way
below $L_2$ for sandstones. Also, $L_2$ suggests that computations with
correlations lengths beyond where lineal path goes to zero are not very
informative. For very irregular pores on 2D SEM images, $P$ and $p$ functions
look much less similar compared to the ceramics case on
\cref{fig:real-xct-cfs}. For 2D images it is also visually easy to analyze full
CF maps, as presented on \cref{fig:real-sem-map} for Sandstone 1 sample. Full
correlation maps show expanded version of directional functions -- asymmetry is
immediately evidenced from $S_2$, $F_{ss}$ and $F_{sv}$ maps. Moreover, the map
for $C_2$ shows higher connectivity in y-direction as compared to x-axis. The
interpretation of full correlation maps is less straightforward, but anisotropy
patterns in directions different from those presented in \cref{fig:directions}
can be spotted.
\begin{figure*}[pt]
  \centering
  \includegraphics[width=0.45\linewidth]{s2-sem.png}
  \hfill
  \includegraphics[width=0.45\linewidth]{c2-sem.png}
  \vskip\baselineskip
  \includegraphics[width=0.45\linewidth]{l2-sem.png}
  \vskip\baselineskip
  \includegraphics[width=0.45\linewidth]{surf2-sem.png}
  \hfill
  \includegraphics[width=0.45\linewidth]{surfvoid-sem.png}
  \vskip\baselineskip
  \includegraphics[width=0.45\linewidth]{pore-size-sem.png}
  \hfill
  \includegraphics[width=0.45\linewidth]{chord-length-sem.png}
  \caption[]{Plots of correlation functions for SEM samples on
    \cref{fig:real-data}}
  \label{fig:real-sem-cfs}
\end{figure*}

\begin{figure*}[pt]
  \centering
  \subfigure[$S_2$]{
    \includegraphics[width=0.45\linewidth]{map-s2.png}}
  \hfill
  \subfigure[$C_2$]{
    \includegraphics[width=0.45\linewidth]{map-c2.png}}
  \vskip\baselineskip
  \subfigure[$F_{ss}$]{
    \includegraphics[width=0.45\linewidth]{map-surf2.png}}
  \hfill
  \subfigure[$F_{sv}$]{
    \includegraphics[width=0.45\linewidth]{map-surfvoid.png}}
  \caption[]{Correlation maps for Sandstone 1.}
  \label{fig:real-sem-map}
\end{figure*}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.45\linewidth]{multiphase.png}
  \includegraphics[width=0.45\linewidth]{multiphase-cc.png}
  \caption[]{Cross-correlation functions for multiphase sample (soil) which has
    4 distinct phases, labeled as 1-4. Each function $\rho_{ij}$ is normalized
    by division by $\rho_{ii}(0) \rho_{jj}(0)$.}
  \label{fig:real-multiphase}
\end{figure}

@@

\begin{figure*}[t]
  \centering
  \subfigure[Two-dimensional case, module \code{Directional}]{
    \includegraphics[width=0.475\linewidth]{time-2d.png}
    \label{fig:timings-2d}}
  \hfill
  \subfigure[Three-dimensional case, module \code{Directional}]{
    \includegraphics[width=0.475\linewidth]{time-3d.png}
    \label{fig:timings-3d}}
  \vskip\baselineskip
  \subfigure[Two-dimensional case, module \code{Map}, CPU]{
    \includegraphics[width=0.475\linewidth]{time-2d-map.png}
    \label{fig:timings-2d-map}}
  \hfill
  \subfigure[Three-dimensional case, module \code{Map}, CPU]{
    \includegraphics[width=0.475\linewidth]{time-3d-map.png}
    \label{fig:timings-3d-map}}
  \vskip\baselineskip
  \subfigure[Two-dimensional case, module \code{Map}, GPU]{
    \includegraphics[width=0.475\linewidth]{time-2d-gpu.png}
    \label{fig:timings-2d-gpu}}
  \hfill
  \subfigure[Three-dimensional case, module \code{Map}, GPU]{
    \includegraphics[width=0.475\linewidth]{time-3d-gpu.png}
    \label{fig:timings-3d-gpu}}
  \caption[]{Execution times for different correlation functions.}
  \label{fig:timings}
\end{figure*}

@@

\section{Acknowledgments}
This research was supported by the Russian Science Foundation grant 23-74-00061.

Collaborative effort of the authors within the FaT iMP (Flow and Transport in
Media with Pores) research group (www.porenetwork.com) and used some of its
software. We thank our colleague Konstantin Romanenko for suggestions and administrative work.
